---
title: "Getting Started with LOLA"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting Started with LOLA}
output: knitr:::html_vignette
---

# Getting started with LOLA

You need 3 things to run a LOLA analysis:

1. userSets: Regions of interest (at least 1 set of regions, a GRanges object, or multiple sets of regions of interest as a GRangesList object)
2. userUniverse: The set of regions tested for inclusion in your sets of regions of interest (a single GRanges object)
3. regionDB: A database of region sets to test for enrichment of overlap


Let's load an example dataset. The database location should point to a folder that contains collection subfolders:

```{r Load a regionDB}
library("LOLA")
dbPath = system.file("extdata", "hg19", package="LOLA")
regionDB = loadRegionDB(dbPath)
```

The regionDB is an R (list) object that has a few elements: 

```{r}
names(regionDB)
```

* dbLocation: A string recording the location of the database folder you passed to `loadRegionDB()`.
* collectionAnno: A `data.table` annotating the collections, with rows corresponding to the rows in your `collection` annotation files in the database.
* regionAnno: A `data.table` annotating each region set, with rows corresponding to bed files in the database (there is also a `collection` column recording which collection each region set belongs to).
* regionGRL: A `GRangesList` object holding the actual regions, with one list element per region set, ordered as in `regionAnno`.

Now with the database loaded, let's load up some sample data:

```{r Load sample user sets and universe}
data("sample_input", package="LOLA") # load userSet
data("sample_universe", package="LOLA") # load userUniverse
```

Now we have a GRanges object called `userSet` and a GRanges object called `userUniverse`. This is all we need to run the enrichment calculation:

```{r Run the calculation}
locResults = calcLocEnrichment(userSet, userUniverse, regionDB, dbTitle="dbTitle", cores=1)
```

calcLocEnrichment tests for pairwise overlap between each user set and each region set in regionDB. It then uses a Fisher's exact test to assess signficance of the overlap. The results are a `data.table` with several columns:

```{r}
colnames(locResults)
head(locResults)
```

Columns `userSet` and `dbSet` are indexes into the respective GRangeList objects, identifying each pairwise comparison. There are a series of columns describing the results of the statistical test, such as `pValueLog`, `logOdds`, and the actual values from the contingency table (`support` is the overlap, and `b`, `c`, and `d` complete the 2x2 table). Rank columns simply rank the tests by `pValueLog`, `logOdds`, or `support`; following these are a series of columns annotating the database regions, depending on how you populated the `index` table in the regionDB folder.

You can explore these results in R by, for example, ranking with different orders:

```{r}
locResults[order(support, decreasing=TRUE),]
```

And finally, record the results to file like this:

4. Write out results:
```{r Write results}
writeCombinedEnrichment(locResults, outFolder= "lolaResults", includeSplits=TRUE);
```
