---
title: "Getting Started with LOLA"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting Started with LOLA}
output: knitr:::html_vignette
---

# Getting started with LOLA

You need 3 things to run a LOL analysis:

1. userSets: Regions of interest (at least 1 set of regions, a GRanges object, or multiple sets of regions of interest as a GRangesList object)
2. userUniverse: The set of regions tested for inclusion in your sets of regions of interest (a single GRanges object)
3. regionDB: A database of region sets to test for enrichment of overlap


Let's load an example dataset:

```{r Load a regionDB}
library("LOLA")
dbPath = system.file("extdata", "hg19", package="LOLA")
regionDB = loadRegionDB(dbLocation= dbPath)
```

Load up some sample data

```{r Load sample user sets and universe}
data("sample_input", package="LOLA")
data("sample_universe", package="LOLA")
```

Run the enrichment calculation:

```{r Run the calculation}
locResults = calcLocEnrichment(userSet, userUniverse, regionDB, dbTitle="dbTitle", cores=1)
```

4. Write out results:
```
writeCombinedEnrichment(locResults, outFolder= "locationResults", includeSplits=TRUE);
```
